Musical elements in the discrete-time representation of sound

### 2017-08-12 article/acm/body.tex
finished full review of the article

\subsection{Serialism and post-serial techniques}\label{sec:ser}
Expanding the concepts in Section~\ref{sec:atonal},
sequences of characteristics can be predefined and
used throughout a musical piece.
These sequences can be of intensities, timbre, durations,
density of events, etc.
The sequences can be used very strictly or loose,
such as by skipping some elements.
The sequences can be of different sizes, yielding arcs
until the initial condition is reached again (i.e. cycles).
One paradigmatic case is the ``total serialism'' where all
the musical characteristics are serialized.
Although the use of sequences is inherent to music (e.g. scales, metric pulses),
their use with greater emphasis than tonal (or modal) elements
in western music, as an artistic trend, took place only in the
first half of the twentieth century and is called ``serialism''.
Post-serial techniques are numerous, but here is a description of
important concepts:
\begin{itemize}
	\item Spectralism:
		consists on use the (Fourier) spectrum of a sound or the harmonic series for music composition,
		such as
		to derive harmonies, sequences of pitches or a temporal evolution of the overall spectrum~\cite{grisey}.
		In other words, the most prominent frequencies can be used as pitches,
		real notes can be used to mimic an original spectrum
		(e.g. use piano notes to mimic the spectrum of a spoken sentence) or portions of the spectrum made to vary~\cite{spectral}.
	\item Spectromorphology:
		can be considered a spectral music (spectralism) theoretical framework~\cite{smalley,schaeffer}
		that examines the relation between sound spectra and their temporal evolution.
		The theory poses e.g. different onsets, continuations and terminations; characteristics of (sonic) ``motion'';
		and spectral density.
	\item Stochastic music:
		the use of random variables to describe musical elements are largely considered in stochastic music~\cite{formalized}.
		In summary, one can use probability distributions for the synthesis of sounds and larger scale musical elements.
		Changes in these distributions or in other characteristics yield the discourse.
	\item Textures:
		sounds can be assembled in terms of a ``sonic texture''.
		Here, a texture is often thought about very abstractly
		as a sonic counterpart of visual texture.
		Parameters that can be used for such are: range between highest and lowest note,
		density of notes, durations of notes, motives, number of voices, etc.
		If the sounds are small enough (typically < $100ms$) the process can be though of in terms of \emph{granular synthesis}~\cite{microsound}.
\end{itemize}

\subsection{Musical idiom?}
In numerous studies, there are models, discussions and exploitation of a `musical language'.
Some of them are linguistic theories applied to music
and some discern different `musical
idioms'~\cite{Lerdahl, Harmonia, Salzer,Alfaix}. Simply put, a musical idiom
is the result of chosen materials together with variation techniques and
relations established between elements along a music piece. In these matters,
dichotomies are prominent, as explained in Section~\ref{subsec:motivos}:
repetition and variation, relaxation and tension, stability and instability, consonance and dissonance, etc.
A thorough discussion of what can be considered a musical language is out of the scope of this article, but this brief consideration of the subject is useful as a onvergence of all the previous content.

\subsection{Musical usages}\label{subsec:usosmusicais3}
First, the basic note was defined and characterized in quantitative terms in Section~\ref{sec:notaDisc}.
Next, the internal note
composition was addressed within both internal transitions and elementary sonic treatment
(Section~\ref{sec:varInternas}). Finally, this section aims at organizing these notes in music. The numerous resources and consequent infinitude
of praxis possibilities is typical and highly relevant for artistic contexts~\cite{Harmonia,Webern}.

There are studies and further developments for each of the presented resources.
For example, it is possible to obtain `dirty' triadic harmonies (with notes out of the triad) by superposition of perfect fourths.
Another interesting example is the superimposition of rhythms in different metrics, constituting what is
called \emph{polyrhythm}. The music piece \emph{Poli-hit mia} in \massa\ toolbox~\cite{MASSA} explores these simultaneous metrics by impulse trains convolved with notes.

Microtonal scales are important for 20th
century music~\cite{microtonalidade} and yielded diverse remarkable results throughout history, e.g.
fourths of a tone ($\epsilon=2^{\frac{1}{24}}$) is often used in some genres of Indian and Arabic music. The musical sequence \emph{MicroTom} in \massa\ toolbox~\cite{MASSA} explores these resources, including microtonal melodies and harmonies
with many notes in a very reduced note scope.

As in Section~\ref{subsec:mus2}, relations between
parameters are powerful ways to achieve musical pieces.
The number of permuted
notes can vary during the music, revealing relationship with piece
duration. Harmonies can be made from triads (Equations~\ref{triades}) with duplicated
notes at each octave and more numerous duplication when the depth and frequency of
vibratos are lower (Equations~\ref{vbrGamma},~\ref{vbrAux},~\ref{vbrF},~\ref{vbrGamma2},~\ref{vbrT}).
Incontestably, the possibilities are very wide, which is evidenced in the wide range of musical pieces and styles.

The symmetries at octave divisions (Equation~\ref{escSim}) and the
symmetries presented as permutations (Table~\ref{tab:change} and
Equations~\ref{eq:groups}) can be used together. In the music piece \emph{3 trios}
this association is done in a systematic way in order to achieve a specific style. This is an instrumental piece, not included as a source
code but available online~\cite{3Trios}.

\emph{PPEPPS} (Pure Python EP: Project Solvent) is an EP (Extended Play) synthesized using
resources presented in this study. With minimal parametrization, the scripts
generate complete musical pieces, allowing easy composition of sets of
music. A simple script of a few lines specifies music delivered as 16 bit
44.1kHz PCM files (WAVE). This facility and technological
arrangement creates aesthetic possibilities for both sharing and education.

\section{Conclusions and further developments}
\label{cap:conclusao}
In our understanding,
this article is effective in relating musical elements to digital audio.
We aimed at achieving a concise presentation of the subject because it involves many knowledge fields, and therefore can very easily blast into thousands of pages.
Some readers might benefit from the text alone, but the \emph{Scripts} in the \massa\ toolbox, where all the equations and concepts are directly and simply implemented as software (in Python),
are very helpful for one to achieve elaborated implementations and deeper understandings.
The scripts include routines that render musical pieces that illustrate the concepts in practical contexts.
This is valuable since art (music) can involve many non-trivial processes and is often deeply glamorized, which results in a nearly unmanageable terrain for a newcomer.
Moreover, this didactic report and the supplied open source scripts should facilitate the use of the proposed framework.
The Supporting Information document holds listings of sections, equations, figures, tables, scripts and other documents.

The possibilities provided by this exposition pour from both the organization of knowledge and the ability to achieve sounds which are extremely true to the models.
For example, one can produce noises with arbitrary resolution of the spectrum and a musical note can be synthesized with the parameters (e.g. of a vibrato) updated sample-by-sample.
Furthermore, software for synthesis and processing of sounds for musical purposes
by standard restrict the bit depth to 16 or 24.
This is achievable in this framework but by standard Python uses more bits per floating point number.
These ``higher fidelity'' characteristics can be crucial
e.g. for psychoacoustic experiments or to generate high quality musical sounds or pieces.
Simply put, it is compelling
for many scientific and artistic purposes.
The didactic potential of the framework is evident when noticed that:
\begin{itemize}
	\item the integrals and derivatives, ubiquitous in continuous signal processing,
are all replaced, in discrete signals, by summations,
which are more intuitive and does not require fluency in calculus.
	\item The equations and concepts are implemented as software which can be easily assembled and inspected.
\end{itemize}
\noindent In fact, this framework was used in a number of contexts, including courses, software implementations and making music~\cite{vimeoLM,vivace,repoDissertacao,dissertacao}.
Such detailed analytical descriptions, together with the computational implementations, have not been covered before in the literature, as far as the authors know, such as testified in the literature review (Appendix G of~\cite{dissertacao}), where books, articles and open software are related to this framework.

The free software license and online availability of the content,
with the respective codes and sonic examples, facilitate collaborations and the generation of sub-products in a co-authorship fashion,
easing new implementations and development of musical pieces.
The scripts can be divided in three groups: implementation of all the equations and topics of music theory; routines for rendering musical pieces that illustrate the concepts; scripts that render the figures of this article and the article itself.

This framework favored the formation of interest groups in topics such as musical creativity and computer music.
In particular, the project \url{labMacambira.sourceforge.net} groups Brazilian and foreign co-workers
in diverse areas that range from digital direct democracy and georeferencing to art and education.
This was only possible because of the usefulness of audiovisual abilities in many contexts, in particular because of the knowledge and mastery condensed in the \massa\ framework.\footnote{There are more than 700 videos, written documents,
 original software applications and contributions in well-known external software (such as Firefox, Scilab, LibreOffice, GEM/Puredata, to name just a few)~\cite{siteLM,wikiLM,vimeoLM}.
Some of these efforts are available online~\cite{dissertacao}.
It is evident that all these contributions are a consequence of more that just \massa, but it is also evident to the authors that \massa\ had a primary role in converging interests and attracting collaborators.}

Future work might include application of these results in artificial intelligence
for the generation of appealing artistic materials. Some psychoacoustic effects were detected,
which needs validation and should be reported, specially with~\cite{quadrosSonoros}.\footnote{The sonic portraits where sent to a public mailing list and the fifth piece was reported by some individuals to induce a state in which noises from the own tongue, teeth and jaw of the individual echoed for some seconds~\cite{quadrosMetarec}.} Other foreseen advances
are: a JavaScript version of the toolbox, better hypermedia deliverables of this framework, user guides
for different goals (e.g. musical composition, psychophysical experiments, sound synthesis, education), creation
of musical pieces and open experiments to be studied with EEG recordings, and further analytical specification of musical
elements in the discrete-time representation of sound as feedback is received from the community.

>>>>

\subsection{Serialism and post-serial techniques}\label{sec:ser}
Expanding the concepts exposed in Section~\ref{sec:atonal},
sequences of characteristics can be predefined and
used throughout a musical piece.
These sequences can be of intensities, timbre, durations,
density of events, etc.
The sequences can be used very strictly or loosely,
such as by arbitrarily skipping some elements.
The sequences can be of different sizes, yielding arcs
until the initial condition is reached again (i.e. cycles).
One paradigmatic case is the ``total serialism'' where all
the musical characteristics are serialized.
Although the use of sequences is inherent to music (e.g. scales, metric pulses),
their use with greater emphasis than tonal (or modal) elements
in western music, as an artistic trend, took place only in the
first half of the twentieth century and is called ``serialism''.
Post-serial techniques are numerous, but here is a description of
important concepts:
\begin{itemize}
	\item Spectralism:
		consists on use the (Fourier) spectrum of a sound or the harmonic series for music composition,
		such as
		to derive harmonies, sequences of pitches or a temporal evolution of the overall spectrum.
		Examples: the most prominent frequencies of a sound can be used as pitches,
		real notes can be used to mimic an original spectrum
		(e.g. use piano notes to mimic the spectrum of a spoken sentence) or portions of the spectrum made to vary.~\cite{grisey}
	\item Spectromorphology:
		can be considered a spectral music (spectralism) theoretical framework
		that examines the relation between sound spectra and their temporal evolution.
		The theory poses e.g. different onsets, continuations and terminations; characteristics of (sonic) ``motion'';
		and spectral density.~\cite{smalley,schaeffer}
	\item Stochastic music:
		the use of random variables to describe musical elements are largely considered in stochastic music.
		In summary, one can use probability distributions for the synthesis of sounds and larger scale musical elements.
		Changes in these distributions or in other characteristics yield the discourse.~\cite{formalized}
	\item Textures:
		sounds can be assembled in terms of a ``sonic texture''.
		Here, a texture is often thought about very abstractly
		as a sonic counterpart of visual texture.
		Parameters that can be used for such are: range between highest and lowest note,
		density of notes, durations of notes, motives, number of voices, etc.
		If the sounds are small enough (typically $< 100ms$)
		the process can be though of in terms of \emph{granular synthesis}~\cite{microsound}.
\end{itemize}

\subsection{Musical idiom?}
In numerous studies, there are models, discussions and exploitation of a `musical language'.
Some of them are linguistic theories applied to music
and some discern different `musical
idioms'~\cite{Lerdahl, Harmonia, Salzer,Alfaix}. Simply put, a musical idiom
is the result of chosen materials together with variation techniques and
relations established between elements along a music piece. In these matters,
dichotomies are prominent, as explained in Section~\ref{subsec:motivos}:
repetition and variation, relaxation and tension, stability and instability, consonance and dissonance, etc.
A thorough discussion of what can be considered a musical language is out of the scope of this article, but this brief consideration of the subject is useful as a convergence of all the previous content.

\subsection{Musical usages}\label{subsec:usosmusicais3}
We strived to supply the reader with a very useful collection of
fundamental concepts to elaborate musical structures
with the content of Sections~\ref{sec:discNote} and~\ref{sec:internalVar}.
The numerous resources and consequent infinitude
of praxis possibilities is typical and highly relevant for artistic contexts~\cite{Harmonia,Webern}.
For example, it is possible to obtain `dirty' triadic harmonies (with notes out of the triad) by superposition of perfect fourths.
Another interesting example is the superimposition of rhythms in different metrics, constituting what is
called \emph{polyrhythm}. The music piece \emph{Poli-hit mia} explores these simultaneous metrics by impulse trains convolved with notes.~\cite{MASSA} 
Microtonal scales are important for 20th
century music~\cite{microtonalidade} and yielded diverse remarkable results throughout history.
E.g.
fourths of a tone ($\epsilon=2^{\frac{1}{24}}$) is often used in some genres of Indian and Arabic music.
The musical sequence \emph{MicroTom} explores microtonal melodies and harmonies.~\cite{MASSA}
As stated in Section~\ref{subsec:mus2}, relations between
parameters are powerful ways to achieve musical pieces.
The number of permuted
notes can vary during the music, revealing relationship with piece
duration. Harmonies can be made from triads (Equations~\ref{triades}) with duplicated
notes at each octave and more numerous duplications entail lower depth and frequency of
vibratos (Equations~\ref{vbrGamma},~\ref{vbrAux},~\ref{vbrF},~\ref{vbrGamma2},~\ref{vbrT}).
Incontestably, the possibilities are very wide, which is evidenced in the wide range of musical pieces and styles.

The symmetries at octave divisions (Equation~\ref{escSim}) and the
symmetries presented as permutations (Table~\ref{tab:change} and
Equations~\ref{eq:groups}) can be used together. In the music piece \emph{3 trios}
this association is done in a systematic way in order to achieve a specific style (or idiom).
This is an instrumental piece, not included as a source
code but available online~\cite{3Trios}.

\emph{PPEPPS} (Pure Python EP: Project Solvent) is an EP (Extended Play) synthesized using
resources presented in this article. With minimal parametrization, the scripts
generate complete musical pieces, allowing easy composition of sets of
music. A simple script of a few lines specifies music delivered as 16 bit,
44.1kHz PCM files (WAVE). This facility and technological
arrangement creates aesthetic possibilities with potential to both
sharing and education.

\section{Conclusions and further developments}
\label{cap:conclusao}
In our understanding,
this article is effective in relating musical elements to digital audio.
We aimed at achieving a concise presentation of the subject because it involves many knowledge fields, and therefore can very easily blast into thousands of pages.
Some readers might benefit from the text alone, but the \emph{scripts} in the \massa\ toolbox, where all the equations and concepts are directly and simply implemented as software (in Python),
are very helpful for one to achieve elaborated implementations and deeper insights.
The scripts include routines that render musical pieces that illustrate the concepts in practical contexts
even when we did not use equations in the descriptions (mainly in Section~\ref{sec:notesMusic}).
This is valuable since art (music) can involve many non-trivial processes and is often deeply glamorized, which results in a nearly unmanageable terrain for a newcomer.
Moreover, this didactic report and the supplied open source scripts should facilitate the use of the proposed framework.
The Supporting Information document holds listings of sections, equations, figures, tables, scripts and other documents.

The possibilities provided by this exposition pour from both the organization of knowledge and the ability to achieve sounds which are extremely true to the models.
For example, one can produce noises with arbitrary resolution of the spectrum and a musical note can be synthesized with the parameters (e.g. of a vibrato) updated sample-by-sample.
Furthermore, software for synthesis and processing of sounds for musical purposes
by standard restrict the bit depth to 16 or 24.
This is achievable in this framework but by standard Python uses more bits per floating point number.
These ``higher fidelity'' characteristics can be crucial
e.g. for psychoacoustic experiments or to generate high quality musical sounds or pieces.
Simply put, it is compelling
for many scientific and artistic purposes.
The didactic potential of the framework is evident when noticed that:
\begin{itemize}
	\item the integrals and derivatives, ubiquitous in continuous signal processing,
are all replaced, in discrete signals, by summations,
which are generally more intuitive and do not require fluency in calculus.
	\item The equations and concepts are implemented as software which can be easily assembled and inspected.
\end{itemize}
\noindent In fact, this framework was used in a number of contexts, including courses, software implementations and making music~\cite{vimeoLM,vivace,repoDissertacao,dissertacao}.
Such detailed analytical descriptions, together with the computational implementations, have not been covered before in the literature, as far as the authors know, such as testified in the literature review (Appendix G of~\cite{dissertacao}), where books, articles and open software are related to this framework.

The free software license and online availability of the content,
with the respective codes and sonic examples, facilitate collaborations and the generation of sub-products in a co-authorship fashion,
favoring new implementations and musical pieces.
The scripts can be divided in three groups: implementation of all the equations and topics of music theory; routines for rendering musical pieces that illustrate the concepts; scripts that render the figures of this article and the article itself.
There are also a few auxiliary scripts which did not fit into these categories.~\cite{SI}

% This framework favored the formation of interest groups in topics such as musical creativity and computer music.
% In particular, the project \url{labMacambira.sourceforge.net} groups Brazilian and foreign co-workers
% in diverse areas that range from digital direct democracy and georeferencing to art and education.
% This was only possible because of the usefulness of audiovisual abilities in many contexts, in particular because of the knowledge and mastery condensed in the \massa\ framework.\footnote{There are more than 700 videos, written documents,
%  original software applications and contributions in well-known external software (such as Firefox, Scilab, LibreOffice, GEM/Puredata, to name just a few)~\cite{siteLM,wikiLM,vimeoLM}.
% Some of these efforts are available online~\cite{dissertacao}.
% It is evident that all these contributions are a consequence of more that just \massa, but it is also evident to the authors that \massa\ had a primary role in converging interests and attracting collaborators.}

Future work might include application of these results in artificial intelligence
for the generation of appealing artistic materials. Some psychoacoustic effects were detected,
which needs validation and should be reported, especially with~\cite{quadrosSonoros}.\footnote{The sonic portraits where sent to a public mailing list and the fifth piece was reported by some individuals to induce a state in which noises from the own tongue, teeth and jaw of the individual echoed for some seconds~\cite{quadrosMetarec}.} Other foreseen advances
are: a JavaScript version of the toolbox, better hypermedia deliverables of this framework, user guides
for different goals (e.g. musical composition, psychophysical experiments, sound synthesis, education), creation
of musical pieces and open experiments to be studied with EEG recordings, and further analytical specification of musical
elements in the discrete-time representation of sound as feedback is received from the community.

### 2017-08-12 article/acm/sample-acmsmall.tex
finished full review of the article

\begin{abstract}
The representation of basic elements of music - such as notes,
ornaments and intervals - in terms of discrete audio
signals is often used in software for music creation and design.
Nevertheless, there is no unified approach that relates these elements to the discrete samples of digitized sound.
In this article, each musical element is related by equations and algorithms to the discrete-time samples of sounds,
and each of these relations are implemented in scripts within a software toolbox,
referred to as \massa\ (Music and Audio in Sample Sequences).
The toolbox also includes this article and every script necessary to render the images
and musical examples, pieces and albums, which adds to the educational and artistic uses.
The fundamental element, the musical note with duration, volume, pitch and timbre,
is related quantitatively to characteristics of the digital signal.
Internal variations of a note, such as tremolos, vibratos and spectral fluctuations,
are also considered, which enables the synthesis of notes inspired by real instruments and new sonorities.
With this representation of notes, resources are provided for the generation of higher scale musical structures,
such as rhythmic meter, pitch intervals and cycles.
This framework enables precise and trustful scientific experiments and is useful for education and art.
The efficacy of \massa\ is confirmed by the synthesis of small musical pieces using basic notes,
elaborated notes and notes in music, which reflects the organization of the toolbox and of this article.
It is possible to synthesize whole albums through collage of the scripts and settings specified by the user.
With the open source paradigm, the toolbox can be promptly scrutinized,
expanded in co-authorship processes and used with freedom by musicians, engineers and other interested parties.
In fact, \massa\ has already been employed by external users for diverse purposes which include music production,
artistic presentations, psychoacoustic experiments and computer language diffusion where the appeal of audiovisual artifacts is exploited for education.
\end{abstract}

>>>>

\begin{abstract}
The representation of basic elements of music - such as notes - in terms of discrete audio
signals is often used in software for music creation and design.
Nevertheless, there is no unified approach that relates these elements to the discrete samples of digitized sound.
In this article, each musical element is related by equations and algorithms to the discrete-time samples of sounds,
and each of these relations are implemented in scripts within the
    \massa\ software toolbox
 (Music and Audio in Sample Sequences).
The toolbox also includes this article and every script necessary to render the images
and musical examples, pieces and albums.
The fundamental element, the musical note with duration, volume, pitch and timbre,
is related quantitatively to characteristics of the digital signal.
Internal variations of a note, such as tremolos, vibratos and spectral fluctuations,
are also described, which enables the synthesis of notes inspired by real instruments and new sonorities.
With this representation of notes, resources are provided for the
    generation of higher level musical structures,
such as rhythmic meter, pitch intervals and cycles.
This framework enables precise and trustful scientific experiments and is useful for education and art.
The efficacy of \massa\ is confirmed by the synthesis of small musical pieces using basic notes,
elaborated notes and notes in music, which reflects the organization of the toolbox and of this article.
It is possible to synthesize whole albums through collage of the scripts and settings specified by the user.
With the open source paradigm, the toolbox can be promptly scrutinized,
expanded in co-authorship processes and used with freedom by musicians, engineers and other interested parties.
\end{abstract}
